### 2020-05-31 case-studies.tex
hygiene section

\subsubsection{Typed Hygienic Splicing and Parameterization} 
Livelit expressions can have sub-expressions provided by the client in one of two ways:
as parameters or as splices.

\paragraph{Parameters} Parameters appear immediately after the livelit name. 
For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
Livelits can declare a fixed number of parameters and specify their types in their declarations. 
For example, the declaration of \li{\$cutoffs} above specifies one parameter of type \li{List(Float)}.
Parameters can be partially applied in livelit abbreviations. For example, consider the slider livelit
from Fig.~\ref{fig:color}:
\begin{lstlisting}[numbers=none]
livelit $slider (min: Float) (max: Float) at Float { ... }
\end{lstlisting}
We can partially apply the first parameter as follows to create a non-negative slider with one remaining parameter:
\begin{lstlisting}[numbers=none]
livelit $nnslider = $slider 0.0 in ...
\end{lstlisting}
The client must provide all remaining parameters to use a livelit to fill a hole in the program.
So writing \li{\$nnslider} in expression position will not display the slider GUI. Instead, it will display as a livelit parameterization error.%
\footnote{In Hazel, erroneous expressions are placed inside holes so that they do not prevent other parts of the program from executing
\cite{HazelnutLive}.}

We can apply both parameters in an abbreviation to create an unparameterized livelit:
\begin{lstlisting}[numbers=none]
livelit $percentage = $slider 0.0 1.0 in ...
\end{lstlisting}

\paragraph{Splices} Splices appear directly inside the livelit GUI.
For example, the column headings and the cells in the \li{\$}

unlike arguments, not a fixed number of splices (e.g. adding a new row or column by pressing those buttons)

\subsubsection{Live Evaluation} abc

>>>>

\subsubsection{Typed Hygienic Splicing and Parameterization}\label{sec:splicing-and-parameterization}
Livelit expressions can include sub-expressions in one of two ways:
as parameters or as splices.

\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit. 
For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
The livelit declares a fixed number of parameters and specifies their types in its declaration. 
For example, the declaration of \li{\$cutoffs} shown above specifies one parameter of type \li{List(Float)}
(the grades data to be plotted, see below).

Livelit abbreviations can partially apply parameters. Consider the slider livelit
from Fig.~\ref{fig:color}:
\begin{lstlisting}[numbers=none]
livelit $slider (min: Float) (max: Float) at Float { ... }
\end{lstlisting}
We can partially apply the first parameter as follows to define a parameterized non-negative slider livelit with one remaining parameter:
\begin{lstlisting}[numbers=none]
livelit $nnslider = $slider 0.0 in ...
\end{lstlisting}
Only livelits with no remaining parameters can be used to fill a hole.
So writing \li{\$nnslider} in expression position will not display the slider GUI. Instead, it will display as a ``missing livelit parameter'' error.%
\footnote{\label{footnote:typing}In Hazel, erroneous expressions are automatically placed inside holes so that they do not prevent other parts of the program from executing
\cite{HazelnutLive}.}

\paragraph{Splices}\label{sec:splices}
Spliced expressions (or \emph{splices}) appear directly inside the livelit GUI.
Splices can be filled with Hazel expressions of any form, including other livelit expressions, 
and all of Hazel's editing affordances are available when editing these expressions.
For example, each cell in the \li{\$dataframe} GUI in Fig.~\ref{fig:grading} 
has a corresponding splice, which appears in the formula bar at the top of the livelit 
when the user has selected that cell. 
The cell itself displays the live value of the spliced expression, 
following the example of spreadsheet interfaces 
(we return to live evaluation in Sec.~\ref{sec:live-evaluation} below).

The livelit provides an expected type for each splice.
For example, the splices for the row and column keys in Fig.~\ref{fig:grading} are expected to be expressions
of type \li{String}, and the remaining cells are expected to be expressions of type \li{Float}. 
Hazel surfaces this typing information for the programmer when the cursor is in a splice, 
so it is not necessary to inspect the expansion or the livelit implementation.
If an expression of invalid type is entered, it will display in an error hole as usual,
and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).

Unlike parameters, the number of splices is not fixed in the livelit declaration. Splices can be created, 
deleted, and filled through user interaction with the livelit. For example, clicking the \li{+} buttons
in Fig.~\ref{fig:grading} will create new rows or columns, which will in turn generate new splices.

\paragraph{Hygiene}\label{sec:hygiene}
Parameters and splices can both appear in the underlying expansion. For example,
the expansion of the \li{\$dataframe} livelit in Fig.~\ref{fig:grading} includes the 
spliced expressions that the client entered the cell contents and the row and column keys.
(Not every parameter or splice needs to appear in the expansion. For example,
the bounds provided to the \li{\$slider} livelit are used only to constrain the user 
interface, and do not appear directly in the expansion, which is simply a number literal.)

The client cannot know, without looking at the expansion or the livelit implementation, 
where in the expansion each parameter or splice will appear. This becomes relevant when 
the expansion places the parameter or splice under a binder, e.g. in the body of a function or let binding.
Na\"ively, this could cause inadvertent capture of the bound variable by a free variable 
in the parameter or splice. For example, consider a livelit that generates an expansion 
of the following form:
\begin{lstlisting}[numbers=none]
let len = strlen <splice1> in 
(<splice2>, <splice2> + 1)
\end{lstlisting}
Here, \li{<splice2>} appears under the binding of \li{len}. If the client enters an expression
for \li{<splice2>} that itself refers to a client-side binding of the same variable, \li{len},
it would na\"ively be captured. This sort of bug would difficult to debug, both for the provider
and the client.

To avoid this situation, the livelit expansion mechanism 
maintains parameter and splice hygiene: parameters and splices are included in the expansion 
in a capture-avoiding manner.
Consequently, clients can reason abstractly about binding in splices: variables in splices
always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
(We discuss potentially relaxed variations of this discipline in Sec.~\ref{sec:discussion}).
\subsubsection{Live Evaluation}\label{sec:live-evaluation}
